Bayesian Evidence Synthesis:opioid crisis

### 2020-04-04 latex/newpaper.tex
latex double checked before asking questions

One difficulty in coping with the opioid crisis is that the total number of overdoses is unknown. The reason is that the use of prescription opioids often leads to use of illicit opioids and the extend of  illicit usage cannot be known. Many of those who become addicted to opioids do so after initially receiving a prescription. The highly addictive nature of these pain relievers makes it easy for the human brain to crave more. It is only after their prescription ends that many users realize they’ve become dependent on the effects of opioids  to function “normally.” At that point, they are either forced to get clean and endure the pain that comes with the withdrawal symptoms of opioids or look for another means of getting their high. This is often the time where people will turn to illicit drugs or other analogues. Because prescription opioids are so expensive,  many users turn to heroin. It is often cheaper, more potent, and easier to locate than what they were taking before. In fact, about 80\% of people using heroin started with a prescription to another opioid. After using heroin, however, 23\% of individuals develop opioid addiction.\cite{opioid_desc} \\

@@

\normalsize
[\textbf{phrase differently: the prior and the data relate to all the unknown parameter, but as a last stage you focus on one target.}]
The number of overdoses is our ultimate target of estimation. To achieve the goal in context of Bayesian statistics, we use our prior belief and available data set about the target of estimation. The posterior belief is coming from both of these sources of the information. The following equation express the idea in a mathmatical form:
\begin{equation}
\label{bayes_thm}
\left.\begin{aligned}
p(O_t| y ) = \frac{ p(O_t, y)   }{ p(y)} = \frac{ p(O_t) p(y| O_t)}{ p(y)} \propto p(O_t) p(y| O_t) \end{aligned}\right.
\end{equation}
[\textbf{Question: to make this flow better, maybe you want a graph figure version of (1) which generically describes parameter and latent varaibles given observed data.}]
where $O_t$ is the total number of overdoses in a given month $t$ in British Columbia, and $y$ represents the collected data set providing relevant samples of the target variable. In Bayesian concepts, the variable of interest $O_t$ is a fixed number but the number is yielded from a certain distribution where the distribution describes knowledge about $O_t$; That is, $O_t$ is considered to be a sample of a random variable. The problem, however, is that the data $y$ is not available. Hence it is needed to approach the estimation in an indirect fashion. \\ 
\begin{figure}[htb]
	\centering
	\includegraphics[width=0.5\linewidth]{Figures/simple_drawing}
	\caption{Example of estimating the total overdoses indirectly given two data sets: ambulance attended overdoses and survey data.}
	\label{simple_draw}
\end{figure}

[\textbf{attempted: need more explicitness
	t:month, considering T months all together}]
Figure \ref{simple_draw} shows an example of how the estimation can be proceeded. It is assumed that we are interested in the opioid crisis from month 1 to month T; subscript 1:T for any variable denotes the variable of every target months and subscription t denotes month t. The simplest conceptual model described in Figure \ref{simple_draw} is to take an underlying log-rate $z_t$ where that is independent and identically distributed across months according to a normal distribution with mean $\mu$ and variance $\sigma^2$. \cite{Irvine:modelling} Then let $\lambda_{t}^{OD}$ denote the rate of overdose at month $t$. It is assumed that the number of overdoses in the month, $O_t$, follows Poisson distribution with parameter $\lambda_{t}^{OD}N$ where the population of the region of interest is $N$. Note that the population size, N, is unknown and vary over time in reality but it is assumed known and constant for the investigator here for simplicity.

@@

Since estimation of $O_t$ is not straightforward as some of the variables ($\mu$, $\sigma$) determining $O_t$ is not known, we assume there was a survey conducted to help to support inference about  $O_t$. The survey asked witnesses of opioid overdoses whether someone called an ambulance. This provides an estimate of the proportion of ambulance call $p_A$ among the overdoses. For the moment $p_A$ is assumed constant across time for simplicity. Let $n_{A}$ be the sample size of the survey, and $x_{A}$ be the total number who confirmed that an ambulance was called. It is assumed that $x_{A}$ follows a Binomial distribution:

\begin{equation}
\label{ambulance}
\left.\begin{aligned}
x_{A} \sim Bin(n_{A},p_{A})\end{aligned}\right.
\text{ambulance call-outs model}
\end{equation}

Let \(U_t\) be the number of  ambulance-attended overdoses at a time point $t$. In general, the data of ambulance-attended overdoses \(U_t\) can be obtained. It is assumed that  \(U_t\) follows a Binomial distribution: 
\begin{equation}
\label{over_amb}
\left.
U_t \sim Bin(O_t, p_A)
\right.
\end{equation}

Note that now $O_t$ can be estimated because $p_A$ can be infered from the survey data and the data regarding $U_t$ is given. We suggest a simple model as a start where the model only combines Overdose Model (\ref{overdose}) and Ambulance Call-outs Model (\ref{ambulance}) with the relation (\ref{over_amb}). The next step is to run some simulations to figure out how different types of inputs lead some changes of output. \\

@@

\normalsize 
There exist two data sets: the survey data ($n_A, x_A$), and the ambulance attended overdose data ($U_{1:T}$). The two data sets were simulated as follows. It was assumed that the two data sets were collected for a year (T=12). In terms of the survey data, it is assumed that  we have 12 survey data sets where each survey data set was collected for each month . The true value of $p_A$ was set as $p_A=0.8$ and the survey sample size $n_A$ was set as $n_A=1000$ for every months. The $x_A$ values for each month were independentally generated from the Binomial distribution (\ref{ambulance}).  In terms of the overdose data, it is assumed that the true values of parameters for overdose model were $\mu=\log0.05,\ \sigma=1$.
The vector of $O_t$ was generated following the overdose model (\ref{overdose}). The vector of $U_t$ was generated from the Binomial relation of the two variables ($\ref{over_amb}$). The two generated vectors have the same length with the survey data (T=12).  Note that only $U_t$ and $x_A$ are the only known as the data set in real world and $p_A$ needs to be estimated first in order to estimate $O_t$ which is the ultimate interest of the research.\\

@@

Let $O_T$ denote the total overdose per year ($O_T = O_1 + O_2 + O_3 + ... + O_{12}$). Figure \ref{hst_ot} is the histogram of posterior samples of total overdoses $O_T$ per year. 2000 samples were collected and shown as the histogram.  It is seen that the plot is bell shaped and not skewed. The total number of overdoses per year is inferred to be around 10900 to 11000. [mention about credible interval] [\textbf{not on density scale (count only has meaning relative to chosen MCMC number iterations))}]\\

@@

\normalsize 
Figure \ref{contam_ot} is the boxplots of posterior samples where $p_A$ is contaminated from the survey data. It is seen that the underestimation of $p_A$ (\ref{under_ot}) leads to an overestimation of $O_t$ as the boxplots are above the red dots. This is explainable considering the given data sets (likelihoods) and the relationship between the two models (\ref{over_amb}); given the likelihood of the ambulance attended overdoses, $U_t$, a value of $O_t$ is an unknown parameter of the binomial distribution, $Bin(O_t, p_A)$. Let $o_t$, $u_t$ are posterior samples from $O_t$ and the observed value of $U_t$ respectively. Then $o_t$ is proportional to the quantity generated by multiplying $u_t$ and the inverse of $\hat{p}_{A}$ where $\hat{p}_{A}$ is underestimated estimator of $p_A$; hence $\frac{1}{\hat{p}_{A}}$ is overestimated which leads overestimation of $O_t$:
\begin{equation}
\label{ot.how.made}
\begin{aligned}
o_t \propto u_t \frac{1}{\hat{p}_{A}}
\end{aligned}
\end{equation}
Figure \ref{over_ot} shows the opposite case. Overestimation of $p_A$ leads underestimation of the inverse of $p_A$ which causes underestimation of $O_t$. From both figures it is seen that the bias increases as the estimated values and the actual values get large.\\

@@

Figure \ref{contam_ut} gives the boxplots of posterior predictive samples where $p_A$ is underestimated and overestimated respectively from the survey data. It is seen that none of the contaminations of $p_A$ leads to a an contamination effect on $U_t$ but only affects the $O_t$ estimation.\\

>>>>

One difficulty in coping with the opioid crisis is that the total number of overdoses is unknown. The reason is that the use of prescription opioids often leads to use of illicit opioids and the extent of illicit usage cannot be known. Many of those who become addicted to opioids do so after initially receiving a prescription. The highly addictive nature of these pain relievers makes it easy for the human brain to crave more. It is only after their prescription ends that many users realize they’ve become dependent on the effects of opioids  to function “normally.” At that point, they are either forced to get clean and endure the pain that comes with the withdrawal symptoms of opioids or look for another means of getting their high. This is often the time where people will turn to illicit drugs or other analogues. Because prescription opioids are so expensive,  many users turn to heroin. It is often cheaper, more potent, and easier to locate than what they were taking before. In fact, about 80\% of people using heroin started with a prescription to another opioid. After using heroin, however, 23\% of individuals develop opioid addiction.\cite{opioid_desc} \\

@@

\normalsize
[\textbf{phrase differently: the prior and the data relate to "all the unknown parameter", but as a last stage you focus on one target.}]
The number of overdoses is our ultimate target of estimation. To achieve the goal in context of Bayesian statistics, we use our prior belief and available data set about the target of estimation. The posterior belief is coming from both of these sources of the information. The following equation express the idea in a mathmatical form:
\begin{equation}
\label{bayes_thm}
\left.\begin{aligned}
p(O_t| y ) = \frac{ p(O_t, y)   }{ p(y)} = \frac{ p(O_t) p(y| O_t)}{ p(y)} \propto p(O_t) p(y| O_t) \end{aligned}\right.
\end{equation}
[\textbf{Question: to make this flow better, maybe you want a graph figure version of (1) which generically describes parameter and latent varaibles given observed data.}]
where $O_t$ is the total number of overdoses in a given month $t$ in British Columbia, and $y$ represents the collected data set providing relevant samples of the target variable. In Bayesian concepts, the variable of interest $O_t$ is a fixed number but the number is yielded from a certain distribution where the distribution describes knowledge about $O_t$; That is, $O_t$ is considered to be a sample of a random variable. The problem, however, is that the data $y$ is not available. Hence it is needed to approach the estimation in an indirect fashion. \\ 
\begin{figure}[htb]
	\centering
	\includegraphics[width=0.5\linewidth]{Figures/simple_drawing}
	\caption{Example of estimating the total overdoses indirectly given two data sets: ambulance attended overdoses and survey data.}
	\label{simple_draw}
\end{figure}

Figure \ref{simple_draw} shows an example of how the estimation can be proceeded. It is assumed that we are interested in the opioid crisis from month 1 to month T; subscript 1:T for any variable denotes the variable of every target months and subscription t denotes month t. The simplest conceptual model described in Figure \ref{simple_draw} is to take an underlying log-rate $z_t$ where that is independent and identically distributed across months according to a normal distribution with mean $\mu$ and variance $\sigma^2$. \cite{Irvine:modelling} Then let $\lambda_{t}^{OD}$ denote the rate of overdose at month $t$. It is assumed that the number of overdoses in the month, $O_t$, follows Poisson distribution with parameter $\lambda_{t}^{OD}N$ where the population of the region of interest is $N$. Note that the population size, N, is unknown and vary over time in reality but it is assumed known and constant for the investigator here for simplicity.

@@

Since estimation of $O_t$ is not straightforward as some of the variables ($\mu$, $\sigma$) determining $O_t$ is not known, we assume there was a survey conducted to help to support the inference about  $O_t$. The survey asked witnesses of opioid overdoses whether someone called an ambulance for each month. This provides an estimate of the proportion of ambulance call $p_{A,t}$ among the overdoses. For the moment $p_{A,t}$ is assumed constant across time for simplicity. Let $n_{A,t}$ be the sample size of the survey, and $x_{A,t}$ be the total number who confirmed that an ambulance was called. It is assumed that $x_{A}$ follows a Binomial distribution:

\begin{equation}
\label{ambulance}
\left.\begin{aligned}
x_{A,t} \sim Bin(n_{A,t},p_{A,t})\end{aligned}\right.
\text{ambulance call-outs model}
\end{equation}

Let \(U_t\) be the number of  ambulance-attended overdoses at a time point $t$. In general, the data of ambulance-attended overdoses \(U_t\) can be obtained. It is assumed that  \(U_t\) follows a Binomial distribution: 
\begin{equation}
\label{over_amb}
\left.
U_t \sim Bin(O_t, p_{A,t})
\right.
\end{equation}

Note that now $O_t$ can be estimated because $p_{A,t}$ can be infered from the survey data and the data regarding $U_t$ is given. We suggest a simple model as a start where the model only combines Overdose Model (\ref{overdose}) and Ambulance Call-outs Model (\ref{ambulance}) with the relation (\ref{over_amb}). The next step is to run some simulations to figure out how different types of inputs lead some changes of output. \\

@@

\normalsize 
There exist two data sets: the survey data ($n_{A,1:T}, x_{A,1:T}$), and the ambulance attended overdose data ($U_{1:T}$). The two data sets were simulated as follows. It was assumed that the two data sets were collected for a year (T=12). In terms of the survey data, it is assumed that  we have 12 survey data sets where each survey data set was collected for each month . The true value of $p_{A,t}$ was set as $p_{A,t}=0.8$ and the survey sample size $n_{A,t}$ was set as $n_{A,t}=1000$ for every months (t=1:T). The $x_{A,t}$ values for each month were independentally generated from the Binomial distribution (\ref{ambulance}).  In terms of the overdose data, it is assumed that the true values of parameters for overdose model were $\mu=\log0.05,\ \sigma=1$.
The vector of $O_t$ was generated following the overdose model (\ref{overdose}). The vector of $U_t$ was generated from the Binomial relation of the two variables ($\ref{over_amb}$). The two generated vectors have the same length with the survey data (T=12).  Note that only $U_t$ and $x_{A,t}$ are the only known as the data set in real world and $p_{A,t}$ needs to be estimated first in order to estimate $O_t$ which is the ultimate interest of the research.\\

@@

Let $O_T$ denote the total overdose per year ($O_T = \sum_{t=1}^{12}O_t $). Figure \ref{hst_ot} is the histogram of posterior samples of the total number of overdoses $O_T$ per year. 2000 samples were collected and shown as the histogram.  It is seen that the plot is bell shaped and not skewed. The total number of overdoses per year is inferred to be around 10900 to 11000. [mention about credible interval] [\textbf{not on density scale (count only has meaning relative to chosen MCMC number iterations))}]\\

@@

\normalsize 
Figure \ref{contam_ot} is the boxplots of posterior samples where $p_A$ is contaminated from the survey data. It is seen that the underestimation of $p_A$ (\ref{under_ot}) leads to an overestimation of $O_t$ as the boxplots are above the red dots. This is explainable considering the given data sets (likelihoods) and the relationship between the two models (\ref{over_amb}); given the likelihood of the ambulance attended overdoses, $U_t$, a value of $O_t$ is an unknown parameter of the binomial distribution, $Bin(O_t, p_A)$. Let $o_t$, $u_t$ denote posterior samples from $O_t$ and the observed value of $U_t$ respectively. Then $o_t$ is proportional to the quantity generated by multiplying $u_t$ and the inverse of $\hat{p}_{A}$ where $\hat{p}_{A}$ is underestimated estimator of $p_A$; hence $\frac{1}{\hat{p}_{A}}$ is overestimated which leads overestimation of $O_t$:
\begin{equation}
\label{ot.how.made}
\begin{aligned}
o_t \propto u_t \frac{1}{\hat{p}_{A}}
\end{aligned}
\end{equation}
Figure \ref{over_ot} shows the opposite case. Overestimation of $p_A$ leads underestimation of the inverse of $p_A$ which causes underestimation of $O_t$. From both figures it is seen that the bias increases as the estimated values and the actual values get large.\\

@@

Figure \ref{contam_ut} gives the boxplots of posterior predictive samples where $p_A$ is underestimated and overestimated respectively from the survey data. It is seen that none of the contaminations of $p_A$ leads to an contamination effect on $U_t$ but only affects the $O_t$ estimation.\\
